---
pagetitle: "COVID-19 surveillance summary"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse) ; library(lubridate) ; library(scales) ; library(shiny) ; library(reactable)

id <- "Clifford"
# account for 4-day reporting lag
current_period <- as.Date(c("2020-08-28", "2020-09-03"))
previous_period <- as.Date(c("2020-08-21", "2020-08-27"))

# Postcodes
# Source: ONS Open Geography Portal
postcodes <- read_csv("../data/geospatial/postcodes.csv") %>% 
  select(postcode, wd19cd, wd19nm)

# Confirmed cases: official sensitive data
# Source: PHE Covid-19 Situational Awareness Explorer
raw <- read_csv("../data/data.csv") %>% 
  mutate(date = as.Date(`Specimen Date`, format = "%Y-%m-%d")) %>% 
  left_join(postcodes, by = c("Patient Postcode" = "postcode")) 

cases <- filter(raw, wd19nm == id, date <= current_period[2])

# Mid-2018 population estimates for Trafford's wards
# Source: Nomis / ONS
# URL: https://www.nomisweb.co.uk/datasets/pestsyoaoa
population <- read_csv("http://www.nomisweb.co.uk/api/v01/dataset/NM_2010_1.data.csv?geography=1660945005...1660945019,1660945021,1660945020,1660945022...1660945025&date=latest&gender=0&c_age=200&measures=20100&select=date_name,geography_name,geography_code,gender_name,c_age_name,measures_name,obs_value,obs_status_name") %>% 
  select(wd19nm = GEOGRAPHY_NAME, n = OBS_VALUE)
```

```{css}
@import url('https://fonts.googleapis.com/css2?family=Lato&display=swap');
.div {
  font-family: 'Lato', sans-serif;
}
.title {
  font-size: 16px;
  font-weight: bold;
}
.subtitle {
  margin: 8px 0;
  font-size: 14px;
}
.caption {
  text-align: right;
  font-size: 12px;
  color: #212121;
}
```

# COVID-19 surveillance summary for `r id`

*This summary of COVID-19 surveillance data for `r id` ward focusses on the 14 days between `r format(previous_period[1], '%d %B')` and `r format(current_period[2], '%A %d %B %Y')`.*

**Disclaimer**: The charts and tables in this summary are derived from data shared by Public Health England with local authorities for the purposes of monitoring the COVID-19 outbreak. Ward level data are not officially published and could contain personally identifiable information so <u>please do not share</u> this summary with third parties.
```{r, summary}
# Weekly cases
current_week <- cases %>% 
  group_by(date) %>% 
  summarise(n = n()) %>% 
  filter(date >= current_period[1])

previous_week <- cases %>% 
  group_by(date) %>% 
  summarise(n = n()) %>% 
  filter(date >= previous_period[1], date <= previous_period[2])

# Median age
median_age <- cases %>% 
  group_by(Pillar) %>% 
  summarise(median_age = as.integer(median(`Age (Years)`)))
```

---

- **`r nrow(cases)`** cases have been confirmed in `r id` since the first case was reported on `r format(min(cases$date), '%A %d %B')` which is equivalent to **`r comma(nrow(cases)/population[population$wd19nm == id, ]$n*100000, accuracy = 1)`** cases per 100,000 residents.
- There were **`r sum(current_week$n)`** confirmed cases in `r id` during the week ending `r format(current_period[2], '%A %d %B')` which is equivalent to **`r round(sum(current_week$n)/population[population$wd19nm == id, ]$n*100000,1)`** cases per 100,000 residents. **`r sum(previous_week$n)`** cases or **`r round(sum(previous_week$n)/population[population$wd19nm == id, ]$n*100000,1)`** cases per 100,000 residents were confirmed during the previous week ending `r format(previous_period[2], '%A %d %B')`.
- The median age of residents who tested positive under Pillar 1 since the beginning of the coronavirus outbreak is **`r pull(filter(median_age, Pillar == "PILLAR 1"),median_age)`** years and **`r pull(filter(median_age, Pillar == "PILLAR 2"),median_age)`** years under Pillar 2.

---

<div class="container-fluid">

### Weekly cases
<div class = "row">
<div class = "col-sm-6">
<div class = "div">
```{r, epidemic curve, dev = 'svg'}
weekly_cases <- cases %>% 
  mutate(week = floor_date(date, "1 week", week_start = 5)) %>% 
  group_by(week) %>% 
  summarise(n = n()) %>% 
  ungroup()

div(class = "subtitle",
    div(class = "title", "Epidemic curve of confirmed cases"),
    paste0(id, ", ", format(min(weekly_cases$week), "%d %B"), " to ", format(current_period[2], "%d %B %Y")))

ggplot() +
  geom_hline(yintercept = 0, size = 0.3, colour = "#333333") +
  geom_col(data = weekly_cases, aes(x = week, y = n), fill = "#39809E", alpha = 0.6) +
  scale_x_date(expand = c(0.005, 0.005), date_labels = "%b") +
  scale_y_continuous(expand = c(0.005, 0.005), breaks = function(x) unique(
    floor(pretty(seq(0, (max(x) + 1) * 1.1)))), position = "right") +
  labs(x = "Specimen date", y = NULL, title = NULL, subtitle = NULL, caption = NULL) +
  theme_minimal() +
  theme(plot.margin = unit(rep(0.5, 4), "cm"),
        panel.background = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.title.x = element_text(size = 10, hjust = 0, margin = margin(t = 10)))

div(class = "caption", p("Source: PHE Covid-19 Situational Awareness Explorer"))
```
</div>
</div>

<div class = "col-sm-6">
<div class = "div">
```{r, pillars}
pillars <- cases %>% 
  mutate(week = floor_date(date, "1 week", week_start = 5)) %>% 
  group_by(week, Pillar) %>% 
  summarise(n = n()) %>% 
  ungroup()

div(class = "subtitle",
    div(class = "title", "Confirmed cases by testing pillar"),
    paste0(id, ", ", format(min(pillars$week), "%d %B"), " to ", format(current_period[2], "%d %B %Y")))

ggplot() +
  geom_hline(yintercept = 0, size = 0.3, colour = "#333333") +
  geom_col(data = pillars, aes(x = week, y = n, fill = Pillar), alpha = 1, position = position_stack(reverse = T)) +
  scale_fill_manual(values = c("PILLAR 1" = "#0A91A2", "PILLAR 2" = "#74C9B9"), 
                    name = NULL, 
                    labels = c("PILLAR 1" = "Pillar 1", "PILLAR 2" = "Pillar 2"),
                    na.translate = FALSE,
                    guide = guide_legend(keyheight = unit(1, units = "mm"),
                                         keywidth = unit(12, units = "mm"),
                                         label.position = "bottom",
                                         nrow = 1)) +
  scale_x_date(expand = c(0.005, 0.005), date_labels = "%b") +
  scale_y_continuous(expand = c(0.005, 0.005), breaks = function(x) unique(
    floor(pretty(seq(0, (max(x) + 1) * 1.1)))), position = "right") +
  labs(x = "Specimen date", y = NULL, title = NULL, subtitle = NULL, caption = NULL) +
  theme_minimal() +
  theme(plot.margin = unit(rep(0.5, 4), "cm"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        plot.title.position = "plot",
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12, margin = margin(b = 20)),
        plot.caption = element_text(colour = "grey60", margin = margin(t = 20, b = -10)),
        legend.position = "top", 
        legend.justification = "right",
        legend.direction = "horizontal",
        legend.text = element_text(size = 10),
        axis.title.x = element_text(size = 10, hjust = 0, margin = margin(t = 10)))

div(class = "caption", p("Source: PHE Covid-19 Situational Awareness Explorer"))
```
</div>
</div>
</div>

### Ward comparison
<div class = "row">
<div class = "col-sm-12">
<div class = "div">
```{r, benchmarking}
ward_cases <- raw %>% 
  group_by(wd19nm, date) %>% 
  summarise(n = n()) %>% 
  ungroup() %>% 
  arrange(wd19nm, date) %>% 
  complete(wd19nm = wd19nm,
           date = seq.Date(min(date), max(date), by = "day")) %>% 
  mutate(n = replace_na(n, 0)) %>% 
  ungroup() %>% 
  fill(date) %>% 
  mutate(period = case_when(
    date >= current_period[1] & date <= current_period[2] ~ "current_week",
    date >= previous_period[1] & date <= previous_period[2] ~ "previous_week"
  )) %>% 
  filter(!is.na(period)) %>% 
  select(-date) %>%
  group_by(wd19nm, period) %>% 
  summarise(n = sum(n)) %>% 
  pivot_wider(names_from = period, values_from = n) %>% 
  select(wd19nm, previous_week, current_week) %>% 
  left_join(population, by = "wd19nm") %>% 
  mutate(previous_week_rate = round(previous_week/n*100000,1),
         current_week_rate = round(current_week/n*100000,1),
         change = current_week-previous_week,
         change_rate = round(change/n*100000,1)) %>% 
  ungroup() %>% 
  select(wd19nm, previous_week, current_week, change, previous_week_rate, current_week_rate, change_rate)

ward_table <- reactable(
  ward_cases, 
  defaultPageSize = 7,
  paginationType = "simple",
  resizable = TRUE,
  defaultSorted = "current_week_rate",
  defaultSortOrder = "desc",
  defaultColDef = colDef(align = "left"),
  defaultColGroup = colGroup(align = "left"), 
  rowStyle = function(index) {
    if (ward_cases[index, "wd19nm"] == id) {
      list(background = "rgba(0, 0, 0, 0.05)")
    }
  },
  columns = list(
    wd19nm = colDef(name = "", align = "left"),
    previous_week = colDef(header = paste(format(previous_period[1],'%d %B'),"-", format(previous_period[2],'%d %B')),
                                align = "left"),
    current_week = colDef(header = paste(format(current_period[1],'%d %B'),"-", format(current_period[2],'%d %B')), 
                               align = "left"),
    change = colDef(name = "Change", align = "left"),
    previous_week_rate = colDef(header = paste(format(previous_period[1],'%d %B'),"-", format(previous_period[2],'%d %B')),
                                align = "left"),
    current_week_rate = colDef(header = paste(format(current_period[1],'%d %B'),"-", format(current_period[2],'%d %B')), 
                               align = "left"),
    change_rate = colDef(name = "Change", align = "left")
  ),
  columnGroups = list(
    colGroup(name = "Cases", columns = c("previous_week", "current_week", "change")),
    colGroup(name = "Per 100,00", columns = c("previous_week_rate", "current_week_rate", "change_rate"))
  )
)

div(class = "div",
    div(class = "subtitle",
        div(class = "title", "Weekly confirmed cases per 100,000 residents by ward"),
        paste("Trafford, ", format(previous_period[1], "%d %B"), "to", format(current_period[2], "%d %B %Y"))
        ),
    br(),
    ward_table,
    div(class = "caption", p("Source: PHE Covid-19 Situational Awareness Explorer"))
    )
```

</div>
</div>
</div>

### Age
<div class = "row">
<div class = "col-sm-6">
<div class = "div">
```{r, ageband cumulative, dev = 'svg'}
div(class = "subtitle",
    div(class = "title", "Confirmed cases by age group"),
    paste0(id, ", ", format(min(cases$date), "%d %B"), " to ", format(current_period[2], "%d %B %Y")))

age <- tibble(
  ageband = c("0-4","5-9","10-19","20-29","30-39","40-49","50-59","60-69","70-79","80+")) %>% 
  left_join(
    cases %>%   
      group_by(`Age (Years)`) %>% 
      summarise(n = n()) %>% 
      ungroup() %>% 
      select(age = `Age (Years)`, n) %>% 
      mutate(ageband = cut(age,
                       breaks = c(0,5,10,20,30,40,50,60,70,80,120),
                       labels = c("0-4","5-9","10-19","20-29","30-39","40-49","50-59","60-69","70-79","80+"),
                       right = FALSE)) %>% 
      group_by(ageband) %>% 
      summarise(n = sum(n)) %>% 
      ungroup()
    ) %>% 
  mutate(ageband = fct_relevel(factor(ageband), "0-4","5-9","10-19","20-29","30-39","40-49","50-59","60-69","70-79","80+"),
         n = replace_na(n, 0))

ggplot(data = age) +
  geom_col(aes(x = n, y = ageband), fill = "#F47C54", width = 0.8) +
  geom_vline(xintercept = 0, size = 1, colour = "#333333") +
  geom_label(aes(x = n, y = ageband, label = comma(n, accuracy = 1)),
             hjust = 0,  vjust = 0.5, 
             colour = "#212121", fill = NA, label.size = NA, size = 4) +
  scale_x_continuous(expand = c(0.005, 0.005), limits = c(0, max(age$n)*1.05)) +
  labs(x = NULL, y = NULL, title = NULL, subtitle = NULL, caption = NULL) +
  theme_minimal() +
  theme(plot.margin = unit(rep(0.5, 4), "cm"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 12)) +
  coord_cartesian(clip = "off")

div(class = "caption", p("Source: PHE Covid-19 Situational Awareness Explorer"))
```
</div>
</div>

<div class = "col-sm-6">
<div class = "div">
```{r, ageband fortnight, dev = 'svg'}
div(class = "subtitle",
    div(class = "title", "Confirmed cases by age group"),
    paste0(id, ", 14 days ending ", format(current_period[2], "%d %B %Y")))
 
age <- tibble(
  ageband = c("0-4","5-9","10-19","20-29","30-39","40-49","50-59","60-69","70-79","80+")) %>% 
  left_join(
    cases %>%   
      filter(date >= previous_period[1]) %>% 
      group_by(`Age (Years)`) %>% 
      summarise(n = n()) %>% 
      ungroup() %>% 
      select(age = `Age (Years)`, n) %>% 
      mutate(ageband = cut(age,
                       breaks = c(0,5,10,20,30,40,50,60,70,80,120),
                       labels = c("0-4","5-9","10-19","20-29","30-39","40-49","50-59","60-69","70-79","80+"),
                       right = FALSE)) %>% 
      group_by(ageband) %>% 
      summarise(n = sum(n)) %>% 
      ungroup()
    ) %>% 
  mutate(ageband = fct_relevel(factor(ageband), "0-4","5-9","10-19","20-29","30-39","40-49","50-59","60-69","70-79","80+"),
         n = replace_na(n, 0))

ggplot(data = age) +
  geom_col(aes(x = n, y = ageband), fill = "#F47C54", width = 0.8) +
  geom_vline(xintercept = 0, size = 1, colour = "#333333") +
  geom_label(aes(x = n, y = ageband, label = comma(n, accuracy = 1)),
             hjust = 0,  vjust = 0.5, 
             colour = "#212121", fill = NA, label.size = NA, size = 4) +
  scale_x_continuous(expand = c(0.005, 0.005), limits = c(0, max(age$n)*1.05)) +
  labs(x = NULL, y = NULL, title = NULL, subtitle = NULL, caption = NULL) +
  theme_minimal() +
  theme(plot.margin = unit(rep(0.5, 4), "cm"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 12)) +
  coord_cartesian(clip = "off")

div(class = "caption", p("Source: PHE Covid-19 Situational Awareness Explorer"))
```
</div>
</div>
</div>

### Ethnicity
<div class = "row">
<div class = "col-sm-6">
<div class = "div">
```{r, ethnicity cumulative, dev = 'svg'}
div(class = "subtitle",
    div(class = "title", "Confirmed cases by ethnicity"), paste0(id, ", ", format(min(cases$date), "%d %B"), " to ", format(current_period[2], "%d %B %Y")))

ethnicity <- tibble(
  ethnic_group = c("Asian / Asian British", "Black / Black British", "Mixed", "Other", "White", "Not known")) %>% 
  left_join(
    cases %>% 
      mutate(ethnic_group = 
               case_when(
                 Ethnicity %in% c("British", "Irish", "Any other White background") ~ "White",
                 Ethnicity %in% c("White and Black Caribbean", "White and Black African", "White and Asian", "Any other Mixed / Multiple ethnic background") ~ "Mixed",
                 Ethnicity %in% c("Indian", "Pakistani", "Bangladeshi", "Any other Asian background") ~ "Asian / Asian British",
                 Ethnicity %in% c("Caribbean", "African", "Any other Black / African / Caribbean background") ~ "Black / Black British",
                 Ethnicity %in% c("Chinese", "Any other ethnic group") ~ "Other",
                 TRUE ~ "Not known")) %>% 
      group_by(ethnic_group) %>% 
      summarise(n = n()) %>% 
      ungroup()
    ) %>% 
  mutate(n = replace_na(n, 0))

ggplot(data = ethnicity) +
  geom_col(aes(x = n, y = fct_reorder(ethnic_group, n)), fill = "#F9A953", width = 0.8) +
  geom_vline(xintercept = 0, size = 1, colour = "#333333") +
  geom_label(aes(x = n, y = ethnic_group, label = comma(n, accuracy = 1)),
             hjust = 0,  vjust = 0.5, 
             colour = "#212121", fill = NA, label.size = NA, size = 4) +
  scale_x_continuous(expand = c(0.005, 0.005), limits = c(0, max(ethnicity$n)*1.1)) +
  labs(x = NULL, y = NULL, title = NULL, subtitle = NULL, caption = NULL) +
  theme_minimal() +
  theme(plot.margin = unit(rep(0.5, 4), "cm"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 12)) +
  coord_cartesian(clip = "off")

div(class = "caption", p("Source: PHE Covid-19 Situational Awareness Explorer"))
```
</div>
</div>

<div class = "col-sm-6">
<div class = "div">
```{r, ethnicity fortnight, dev = 'svg'}
div(class = "subtitle",
    div(class = "title", "Confirmed cases by ethnicity"), paste0(id, ", 14 days ending ", format(current_period[2], "%d %B %Y")))

ethnicity <- tibble(
  ethnic_group = c("Asian / Asian British", "Black / Black British", "Mixed", "Other", "White", "Not known")) %>% 
  left_join(
    cases %>% 
      filter(date >= previous_period[1]) %>% 
      mutate(ethnic_group = 
               case_when(
                 Ethnicity %in% c("British", "Irish", "Any other White background") ~ "White",
                 Ethnicity %in% c("White and Black Caribbean", "White and Black African", "White and Asian", "Any other Mixed / Multiple ethnic background") ~ "Mixed",
                 Ethnicity %in% c("Indian", "Pakistani", "Bangladeshi", "Any other Asian background") ~ "Asian / Asian British",
                 Ethnicity %in% c("Caribbean", "African", "Any other Black / African / Caribbean background") ~ "Black / Black British",
                 Ethnicity %in% c("Chinese", "Any other ethnic group") ~ "Other",
                 TRUE ~ "Not known")) %>% 
      group_by(ethnic_group) %>% 
      summarise(n = n()) %>% 
      ungroup()
  ) %>% 
  mutate(n = replace_na(n, 0))

ggplot(data = ethnicity) +
  geom_col(aes(x = n, y = fct_reorder(ethnic_group, n)), fill = "#F9A953", width = 0.8) +
  geom_vline(xintercept = 0, size = 1, colour = "#333333") +
  geom_label(aes(x = n, y = ethnic_group, label = comma(n, accuracy = 1)),
             hjust = 0,  vjust = 0.5, 
             colour = "#212121", fill = NA, label.size = NA, size = 4) +
  scale_x_continuous(expand = c(0.005, 0.005), limits = c(0, max(ethnicity$n)*1.1)) +
  labs(x = NULL, y = NULL, title = NULL, subtitle = NULL, caption = NULL) +
  theme_minimal() +
  theme(plot.margin = unit(rep(0.5, 4), "cm"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 12)) +
  coord_cartesian(clip = "off")

div(class = "caption", p("Source: PHE Covid-19 Situational Awareness Explorer"))
```
</div>
</div>
</div>
</div>
<br />

---

**Glossary of terms**

- **Specimen date**: the date when the sample was taken from the person being tested
- **Epidemic curve**: chart showing the change in the number of cases over time
- **Pillars 1 and 2**: cases reported include those tested in hospital (pillar 1) and in the community (pillar 2)
- **Incidence**: the number of cases per 100,000 residents

**Further resources**

- [COVID-19 page](https://www.trafforddatalab.io/covid19) on the [Trafford Data Lab's](https://www.trafforddatalab.io) website
- [COVID-19 Resources](https://www.mhcctableau.nhs.uk/page/covid-19-resources) developed by [Manchester Health & Care Commissioning](https://www.mhcc.nhs.uk)
- Public Health England's [Coronavirus dashboard](https://coronavirus.data.gov.uk)
- [Our World in Data's](https://ourworldindata.org/coronavirus) coronavirus pages

---

<br /><br />
