---
title: "COVID-19 surveillance summary"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse) ; library(sf) ; library(lubridate) ; library(zoo) ; library(reactable) ; library(classInt) ; library(leaflet) ; library(htmltools) ; library(htmlwidgets) ; library(shadowtext)

## Note that a 4-day lag has been applied to cases retrieved from PHE surveillance systems. This affects the cases data frame.

# UPDATE POSITIVITY RATE!! #
# Source: PHE Daily COVID-19 Surveillance Reports
fortnight_ending <- as.Date("2020-11-29")
positivity <- tribble(
  ~pillar, ~tests, ~test_rate, ~positivity_rate,
  "Pillar 1", NA, NA, NA,
  "Pillar 2", NA, NA, NA,
  "Total", NA, NA, NA) 

# Confirmed cases: official sensitive data
# Source: PHE Covid-19 Situational Awareness Explorer
cases <- read_csv("../data/data.csv") %>% 
  mutate(date = as.Date(`Specimen Date`, format = "%Y-%m-%d")) %>% 
  # add 4-day lag 
  filter(date <= max(date)-days(3))

# Confirmed cases: open data
# Source: PHE Coronavirus Dashboard, Public Health England
# URL: https://coronavirus.data.gov.uk
phe_dashboard <- read_csv("https://coronavirus.data.gov.uk/downloads/csv/coronavirus-cases_latest.csv") %>% 
  filter(`Area type` %in% c("nation", "ltla"),
         `Area name` %in% c("Bolton", "Bury", "Manchester", "Oldham", "Rochdale", "Salford", "Stockport", "Tameside", "Trafford", "Wigan", "England")) %>% 
  mutate(date = as.Date(`Specimen date`, format = "%Y-%m-%d")) %>% 
  select(date,
         area_code = `Area code`,
         area_name = `Area name`,
         new_cases = `Daily lab-confirmed cases`) %>% 
  arrange(date) %>% 
  group_by(area_code, area_name) %>%
  complete(date = seq.Date(min(date), max(date), by = "day")) %>% 
  mutate(new_cases = replace_na(new_cases, 0)) %>% 
  ungroup() %>% 
  fill(area_name) %>%
  bind_rows(
    filter(., area_name != "England") %>% 
      group_by(., date) %>%
      summarise(area_code = "E11000001",
                area_name = "Greater Manchester",
                date = unique(date),
                new_cases = sum(new_cases))
  ) 

# Mid-2019 population estimates
# Source: Nomis / ONS
population <- read_csv("../data/population.csv")

# Mid-2018 population estimates for electoral wards
# Source: Nomis / ONS
ward_population <- st_drop_geometry(st_read("../data/geospatial/wards.geojson")) %>% 
  select(wd19nm, population)

# Trafford residents by age and gender
# Source: Nomis / ONS
ageband <- read_csv("../data/ageband.csv")

# Trafford residents by ethnic group
# Source: 2011 Census, Nomis / ONS
ethnicity <- read_csv("../data/ethnicity.csv")

# Trafford postcodes
# Source: ONS Open Geography Portal
postcodes <- read_csv("../data/geospatial/postcodes.csv")

# Middle-layer Super Output Areas in Trafford
# Source: ONS Open Geography Portal
msoa_boundaries <- st_read("../data/geospatial/msoa.geojson")

# Trafford's electoral wards
ward_boundaries <- st_read("../data/geospatial/wards.geojson")

# Town centres in Trafford
town_centres <- tribble(
  ~name, ~lat, ~lon,
  "Sale", 53.4243665, -2.3182869,
  "Stretford", 53.4452286, -2.3222351,
  "Urmston", 53.4452286, -2.3524475,
  "Altrincham", 53.3838403, -2.3527908,
  "Partington", 53.4212977, -2.4281502,
  "Carrington", 53.433553, -2.384744,
  "Old Trafford", 53.461922, -2.270467,
  "Timperley", 53.3961251, -2.320175)
```

```{css}
@import url('https://fonts.googleapis.com/css2?family=Lato&display=swap');
.div {
  font-family: 'Lato', sans-serif;
}
.title {
  font-size: 16px;
  font-weight: bold;
}
.subtitle {
  margin: 8px 0;
  font-size: 14px;
}
.caption {
  text-align: right;
  font-size: 12px;
  color: #212121;
}
.leaflet-control.legend {
  background: rgba(255,255,255,0.75);
}
```

*This summary draws on a number of Public Health England data sources which have different publication times so the most recent reporting date will vary slightly. A four day lag has been added to the cases retrieved from the PHE Covid-19 Situational Awareness Explorer because of reporting delays.*

<br />
```{r}
# 7-day moving average
moving_average <- cases %>% 
  group_by(date) %>% 
  summarise(n = n()) %>% 
  ungroup() %>% 
  arrange(date) %>% 
  complete(date = seq.Date(min(date), max(date), by = "day")) %>% 
  mutate(n = replace_na(n, 0)) %>% 
  ungroup() %>% 
  fill(date) %>% 
  mutate(seven_day = round(rollmean(n, 7, align = "right", fill = NA),1)) %>%
  filter(date == max(date))

# Weekly cases
current_week <- cases %>% 
  group_by(date) %>% 
  summarise(n = n()) %>% 
  filter(date >= max(date)-days(6))

previous_week <- cases %>% 
  group_by(date) %>% 
  summarise(n = n()) %>% 
  filter(date >= max(date)-days(13), date < max(date)-days(6))

# Pillar
pillar <- cases %>% 
  # last 7 days
  filter(date >= max(date)-days(6) & date <= max(date)) %>% 
  group_by(Pillar) %>% 
  summarise(n = n()) %>% 
  ungroup() %>% 
  mutate(percent = round(n/sum(n)*100,1))
```

---

- The 7-day average number of daily confirmed cases in Trafford on `r format(max(cases$date),format = "%A %d %B %Y")` was **`r pull(moving_average, seven_day)`**.
- There were **`r sum(current_week$n)`** confirmed cases during the week ending `r format(max(current_week$date), '%A %d %B')` which is equivalent to **`r round(sum(current_week$n)/237354*100000,1)`** cases per 100,000 residents. **`r sum(previous_week$n)`** cases or **`r round(sum(previous_week$n)/237354*100000,1)`** cases per 100,000 residents were confirmed during the previous week.
- The highest number of daily cases during the past week were reported on **`r format(pull(select(filter(current_week, n == max(n)), date)), '%A %d %B')`** with **`r max(current_week$n)`** cases.
- **`r paste0(pull(filter(pillar, Pillar == "PILLAR 1"), percent),"%")`** (**`r pull(filter(pillar, Pillar == "PILLAR 1"), n)`**) of cases were tested under Pillar 1 and **`r paste0(pull(filter(pillar, Pillar == "PILLAR 2"), percent),"%")`** (**`r pull(filter(pillar, Pillar == "PILLAR 2"), n)`**) under Pillar 2.

---

<br />
<div class="container-fluid">
<div class = "row">
<div class = "col-sm-7">
<div class = "div">
```{r}
historic_cases <- read_csv("../data/data.csv") %>% 
  mutate(date = as.Date(`Specimen Date`, format = "%Y-%m-%d")) %>% 
  filter(date <= max(date)-days(1)) %>% 
  group_by(date) %>% 
  summarise(n = n()) %>% 
  ungroup() %>% 
  arrange(date) %>% 
  complete(date = seq.Date(min(date), max(date), by = "day")) %>% 
  mutate(n = replace_na(n, 0)) %>% 
  ungroup() %>% 
  fill(date) %>% 
  mutate(ma_cases = rollmean(n, 7, align = "center", fill = NA))

div(class = "subtitle",
    div(class = "title", "Epidemic curve of daily confirmed new cases"),
    paste("Trafford,", format(min(historic_cases$date), "%d %B"), "to", format(max(historic_cases$date), "%d %B %Y")))

ggplot() +
  geom_hline(yintercept = 0, size = 0.3, colour = "#333333") +
  geom_col(data = historic_cases, aes(x = date, y = n), fill = ifelse(historic_cases$date >= max(historic_cases$date)-2, "#bdbdbd", "#39809E"), alpha = 0.6) +
  geom_line(data = historic_cases, aes(x = date, y = ma_cases, colour = "ma_cases"), size = 1) +
  scale_colour_manual(values = c("ma_cases" = "#39809E"), name = NULL, labels = "7-day rolling average") +
  scale_x_date(expand = c(0.005, 0.005), date_labels = "%B") +
  scale_y_continuous(expand = c(0.005, 0.005), breaks = function(x) unique(
    floor(pretty(seq(0, (max(x) + 1) * 1.1)))), position = "right") +
  labs(x = "Specimen date", y = NULL, title = NULL, subtitle = NULL, caption = NULL , tag = "Incomplete cases are shaded in grey") +
  theme_minimal() +
  theme(plot.margin = unit(rep(0.5, 4), "cm"),
        panel.background = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "top", 
        legend.justification = "left",
        legend.direction = "horizontal",
        legend.text = element_text(size = 12),
        axis.title.x = element_text(size = 10, hjust = 0, margin = margin(t = 10)),
        plot.tag.position = c(0.6, 0.9),
        plot.tag = element_text(size = 9, colour = "#757575", hjust = 0),
        axis.text = element_text(size = 12))

div(class = "caption", p("Source: PHE Covid-19 Situational Awareness Explorer"))
```
</div>
</div>

<div class = "col-sm-5">
<div class = "div">
```{r positivity}
positivity_table <- reactable(
  positivity, 
  resizable = TRUE,
  sortable = FALSE,
  defaultColDef = colDef(align = "left"),
  defaultColGroup = colGroup(align = "left"), 
  columns = list(
    pillar = colDef(name = "Pillar"),
    tests = colDef(name = "n", format = colFormat(separators = TRUE)),
    test_rate = colDef(name = "per 100,000", format = colFormat(separators = TRUE)),
    positivity_rate = colDef(name = "Positivity rate")
    ),
  columnGroups = list(
    colGroup(name = "Tests", columns = c("tests", "test_rate"))
  )
  )

div(class = "div",
    div(class = "subtitle",
        div(class = "title", "Testing volume and positivity rate"),
        paste("Trafford, 14 days ending ", format(fortnight_ending, "%d %B"))
        ),
    br(),
    positivity_table,
    div(class = "caption", p("Source: PHE Daily COVID-19 Surveillance Reports"))
    )
```
</div>
</div>
</div>

<div class = "row">
<div class = "col-sm-12">
<div class = "div">
```{r, weekly cases}
weekly_cases <- phe_dashboard %>% 
  mutate(period = case_when(
    date >= max(date)-days(8) & date <= max(date)-days(2) ~ "current_week",
    date >= max(date)-days(15) & date <= max(date)-days(9) ~ "previous_week"
  )) %>% 
  filter(!is.na(period)) %>% 
  select(-date) %>%
  group_by(area_code, area_name, period) %>% 
  summarise(total_cases = sum(new_cases)) %>% 
  pivot_wider(names_from = period, values_from = total_cases) %>% 
  select(area_code, area_name, previous_week, current_week) %>%
  left_join(population, by = "area_code") %>% 
  mutate(previous_week_rate = round(previous_week/population*100000,1),
         current_week_rate = round(current_week/population*100000,1),
         percent_change = (current_week_rate-previous_week_rate)/previous_week_rate) %>% 
  ungroup() %>%
  select(area_name, previous_week_rate, current_week_rate, percent_change) %>% filter(area_name != c("Greater Manchester", "England"))

league_table <- reactable(weekly_cases, 
          defaultPageSize = 12,
          highlight = TRUE, 
          height = "auto",
          compact = TRUE,
          borderless = FALSE,
          wrap = FALSE,
          resizable = TRUE,
          defaultSorted = "percent_change",
          defaultSortOrder = "desc",
          defaultColDef = colDef(align = "left"),
          rowStyle = function(index) {
            if (weekly_cases[index, "current_week_rate"] >= 50) {
              list(background = "rgba(220,24,65,0.1)")
              } else if (weekly_cases[index, "current_week_rate"] >= 25) {
                list(background = "rgba(230,184,63,0.1)")
                } else {
                  list(background = "rgba(40,123,40,0.1)")
                  }
            },
          columns = list(
            area_name = colDef(name = "", sortable = FALSE, align = "left"),
            previous_week_rate = colDef(header = paste(format(max(phe_dashboard$date)-days(15),'%d %b'),"-", format(max(phe_dashboard$date)-days(9),'%d %b')),
                                        align = "left"),
            current_week_rate = colDef(header = paste(format(max(phe_dashboard$date)-days(8),'%d %b'),"-", format(max(phe_dashboard$date)-days(2),'%d %b')), 
                                       align = "left"),
            percent_change = colDef(name = "% change", format = colFormat(percent = TRUE, digits = 1), align = "left"))
          )

div(class = "div",
    div(class = "subtitle",
        div(class = "title", "Weekly incidence per 100,000 population by local authority"),
        paste("Greater Manchester, ", format(max(phe_dashboard$date)-days(15), "%d %B"), "to", format(max(phe_dashboard$date)-days(2), "%d %B %Y"))
        ),
    br(),
    league_table,
    div(class = "caption", p("Source: PHE Coronavirus Dashboard"))
    )
```
</div>
</div>
</div>

<div class = "row">
<div class = "col-sm-12">
<div class = "div">
```{r, local trend, fig.width = 10, dev = 'svg'}
local_trend <- phe_dashboard %>% 
  # exclude incomplete cases
  filter(date <= max(date)-days(2),
         !area_name %in% c("Greater Manchester", "England")) %>% 
  # retain last 10 weeks
  filter(date >= max(date)-weeks(10)) %>%
  mutate(week = floor_date(date, "1 week", week_start = wday(min(date)))) %>% 
  group_by(area_code, area_name, week) %>% 
  summarise(total_cases = sum(new_cases)) %>% 
  ungroup() %>% 
  left_join(population, by = "area_code") %>% 
  mutate(rate_cases = round(total_cases/population*100000,1))

england_cases <- phe_dashboard %>% 
  # exclude incomplete cases
  filter(date <= max(date)-days(2),
         area_name == "England") %>% 
  # retain last 10 weeks
  filter(date >= max(date)-weeks(10)) %>%
  mutate(week = floor_date(date, "1 week", week_start = wday(min(date)))) %>% 
  group_by(area_code, area_name, week) %>% 
  summarise(total_cases = sum(new_cases)) %>% 
  ungroup() %>% 
  left_join(population, by = "area_code") %>% 
  mutate(rate_cases = round(total_cases/population*100000,1))

div(class = "subtitle",
    div(class = "title", "Weekly incidence per 100,000 population by local authority"),
    paste("Greater Manchester,", format(max(phe_dashboard$date-weeks(10)), "%d %B"), "to", format(max(phe_dashboard$date-days(2)), "%d %B %Y")))

ggplot(data = local_trend, aes(x = week, y = rate_cases, group = area_name)) +
  geom_hline(yintercept = 0, size = 1, colour = "#333333") +
  geom_line(colour = "#39809E", size = 0.6) +
  annotate(geom = 'line', x = england_cases$week, y = england_cases$rate_cases, colour = "#bdbdbd",  size = 0.6) +
  scale_x_date(expand = c(0.005, 0.005), date_labels = "%d %b") +
  scale_y_continuous(expand = c(0.005, 0.005), position = "right") +
  labs(x = "Specimen date", y = NULL, title = NULL, subtitle = NULL, 
       caption = NULL, tag = "England = grey line") +
  facet_wrap(~area_name, nrow = 2) +
  theme_minimal() +
  theme(plot.margin = unit(rep(0.5, 4), "cm"),
        panel.background = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.title.x = element_text(size = 10, hjust = 0, margin = margin(t = 10)),
        strip.text = element_text(face = "bold", hjust = 0),
        axis.text.x = element_text(angle = 90, size = 9),
        plot.tag.position = c(0.87, 1.01),
        plot.tag = element_text(size = 9, colour = "#757575", hjust = 0))

div(class = "caption", p("Source: PHE Coronavirus Dashboard"))
```

</div>
</div>
</div>

<div class="row">
<div class = "col-sm-6">
<div class = "div">
```{r, dev = 'svg'}
div(class = "subtitle",
    div(class = "title", "Distribution of positive cases in Trafford"),
    paste("Trafford,", format(max(cases$date)-days(6), "%d %B"), "to", format(max(cases$date), "%d %B %Y")))

rpt_locs <- cases %>% 
  filter(date >= max(date)-days(6)) %>% 
  left_join(postcodes, by = c("Preferred Postcode" = "postcode")) %>% 
  select(wd19nm, lon, lat) %>% 
  group_by(lon, lat) %>%
  summarise(n = n()) %>%
  ungroup() %>%
  arrange(desc(n)) %>% 
  st_as_sf(crs = 4326, coords = c("lon", "lat"))

ggplot() +
  geom_sf(data = ward_boundaries, fill = "#F2F2F2", colour = "#FFFFFF", size = 1) +
  geom_sf(data = rpt_locs, aes(size = n), shape = 21, colour = "#FFFFFF", fill = "#39809E", alpha = 0.5) +
  geom_shadowtext(data = town_centres, aes(x = lon, y = lat, label = name), size = 3.5, colour = "#000000", bg.colour = "#FFFFFF") +
  scale_size_continuous(labels = scales::comma_format(accuracy = 1)) +
  labs(x = NULL, y = NULL, title = NULL, subtitle = NULL, caption = NULL, fill = NULL, size = "Number of cases") +
  coord_sf(datum = NA, expand = FALSE) +
  theme_void() +
  theme(plot.margin = unit(rep(0.5, 4), "cm"),
        legend.title = element_text(face = "bold"),
        legend.text = element_text(size = 12))

div(class = "caption", p("Source: PHE Covid-19 Situational Awareness Explorer | ONS"))
```
</div>
</div>
  
<div class = "col-sm-6">
<div class = "div">
```{r}
# Match MSOA to postcodes -------------------
msoa_cases <- cases %>% 
  filter(date >= max(date)-days(6) & date <= max(date)) %>%
  left_join(postcodes, by = c("Preferred Postcode" = "postcode")) %>% 
  group_by(msoa11cd) %>%
  summarise(n = n()) %>% 
  ungroup() 

# Join cases to msoa boundaries -------------------
msoa <- left_join(msoa_boundaries, msoa_cases, by = "msoa11cd") %>%
  mutate(n = replace_na(n, 0),
         rate = round(n/population*100000,0))

# Set class intervals: natural breaks -------------------
breaks <- classIntervals(msoa[msoa$rate > 0, ]$rate, n = 5, style = "quantile")$brks
pal <- colorBin(palette = "Blues", domain = NULL, bins = breaks, na.color = "#FFFFFF")

# Create label text  -------------------
msoa <- msoa %>% 
  mutate(popup = str_c("<strong>", msoa11hclnm, "</strong><br/>",
                       n, " cases<br/>",
                       rate, " per 100,000") %>% map(HTML))

div(class = "subtitle",
        div(class = "title", "Incidence per 100,000 population by MSOA"),
        paste("Trafford,", format(max(cases$date)-days(6), "%d %B"), "to", format(max(cases$date), "%d %B %Y")))

# Layer up map
bbox <- as.vector(st_bbox(msoa))

leaflet(msoa, height = 350, width = "100%", options = leafletOptions(zoomControl = FALSE, attributionControl = FALSE)) %>% 
  setMaxBounds(bbox[1], bbox[2], bbox[3], bbox[4]) %>%
  addTiles(urlTemplate = "", options = tileOptions(minZoom = 11, maxZoom = 11)) %>%
  addPolygons(fillColor = ~pal(rate), fillOpacity = 0.8, smoothFactor = 0.5, 
              stroke = TRUE, weight = 1, color = "#212121", opacity = 1, 
              label = ~popup, 
              labelOptions = labelOptions(style = list("font-weight" = "normal", padding = "3px 8px"), textsize = "15px", direction = "auto"),
              highlight = highlightOptions(color = "#212121", weight = 2, bringToFront = TRUE)) %>% 
  addLegend(pal = pal, values = ~rate, opacity = 0.8, title = "Rate", position = "bottomright", className = "legend") %>% 
  onRender(
  "function(el, t) {var myMap = this;myMap._container.style['background'] = '#ffffff';}",
        paste0("function(el, x) {$('head').append(","\'<meta name=\"viewport\" content=\"width=device-width, initial-scale=1.0\">\'",");}"))

div(class = "caption", p("Source: PHE Covid-19 Situational Awareness Explorer"))
```
</div>
</div>
</div>

<div class = "row">
<div class = "col-sm-12">
<div class = "div">
```{r}
ward_cases <- cases %>% 
  left_join(postcodes, by = c("Preferred Postcode" = "postcode")) %>% 
  group_by(wd19nm, date) %>% 
  summarise(n = n()) %>% 
  ungroup() %>% 
  arrange(wd19nm, date) %>% 
  complete(wd19nm = wd19nm,
           date = seq.Date(min(date), max(date), by = "day")) %>% 
  mutate(n = replace_na(n, 0)) %>% 
  ungroup() %>% 
  fill(date) %>% 
  mutate(period = case_when(
    date >= max(date)-days(6) & date <= max(date) ~ "current_week",
    date >= max(date)-days(13) & date <= max(date)-days(7) ~ "previous_week"
  )) %>% 
  filter(!is.na(period)) %>% 
  select(-date) %>%
  group_by(wd19nm, period) %>% 
  summarise(n = sum(n)) %>% 
  pivot_wider(names_from = period, values_from = n) %>% 
  select(wd19nm, previous_week, current_week) %>% 
  left_join(ward_population, by = "wd19nm") %>% 
  mutate(previous_week_rate = round(previous_week/population*100000,1),
         current_week_rate = round(current_week/population*100000,1),
         percent_change = (current_week_rate-previous_week_rate)/previous_week_rate) %>% 
  ungroup() %>% 
  select(wd19nm, previous_week_rate, current_week_rate, percent_change)

ward_table <- reactable(
  ward_cases, 
  defaultPageSize = 21,
  paginationType = "simple",
  resizable = TRUE,
  defaultSorted = "percent_change",
  defaultSortOrder = "desc",
  defaultColDef = colDef(align = "left"),
  defaultColGroup = colGroup(align = "left"), 
  columns = list(
    wd19nm = colDef(name = "", align = "left"),
    previous_week_rate = colDef(header = paste(format(max(cases$date)-days(13),'%d %b'),"-", format(max(cases$date)-days(7),'%d %b')),
                                        align = "left"),
    current_week_rate = colDef(header = paste(format(max(cases$date)-days(6),'%d %b'),"-", format(max(cases$date),'%d %b')), 
                                       align = "left"),
   percent_change = colDef(name = "% change", format = colFormat(percent = TRUE, digits = 1), align = "left")
)
)

div(class = "div",
    div(class = "subtitle",
        div(class = "title", "Weekly incidence per 100,000 population by ward"),
        paste("Trafford, ", format(max(cases$date)-days(13), "%d %B"), "to", format(max(cases$date), "%d %B %Y"))
        ),
    br(),
    ward_table,
    div(class = "caption", p("Source: PHE Covid-19 Situational Awareness Explorer | ONS"))
    )
```
</div>
</div>
</div>

<div class = "row">
<div class = "col-sm-6">
<div class = "div">
```{r, dev = 'svg'}
div(class = "subtitle",
    div(class = "title", "Incidence per 100,000 population by age and gender"),
    paste("Trafford,", format(max(cases$date)-days(6), "%d %B"), "to", format(max(cases$date), "%d %B %Y")))

age <- cases %>%   
  # last 7 days
  filter(date >= max(date)-days(6), `Patient Gender` != "Unknown") %>% 
  group_by(`Patient Gender`, `Age (Years)`) %>% 
  summarise(n = n()) %>% 
  ungroup() %>% 
  select(gender = `Patient Gender`, age = `Age (Years)`, n) %>% 
  mutate(ageband = cut(age,
                       breaks = c(0,5,10,20,30,40,50,60,70,80,120),
                       labels = c("0-4","5-9","10-19","20-29","30-39","40-49","50-59","60-69","70-79","80+"),
                       right = FALSE)) %>% 
  group_by(gender, ageband) %>% 
  summarise(n = sum(n)) %>% 
  ungroup() 

# Age-sex pyramid
pyramid <- left_join(ageband, age) %>% 
  mutate(ageband = fct_relevel(factor(ageband), "0-4","5-9","10-19","20-29","30-39","40-49","50-59","60-69","70-79","80+"),
         n = replace_na(n, 0),
         rate = n/population*100000)

spacing = 40
breaks = seq(0, max(pyramid$rate), by = 100)

pyramid_data <- pyramid %>% 
  mutate(rate = case_when(gender == "Male" ~ as.numeric(rate) * -1, TRUE ~ as.numeric(rate)),
         start = if_else(gender == "Male", 0, spacing),
         end = rate + if_else(gender == "Male", 0, spacing),
         mid = (start + end)/2,
         wid = abs(end - start)) 

ggplot(data = pyramid_data) +
  geom_tile(aes(mid, ageband, fill = gender, width = wid), height = 0.8) +
  geom_text(data = . %>% distinct(ageband), aes(spacing/2, ageband, label = ageband)) +
  scale_fill_manual(values = c("#d8b365", "#5ab4ac"), labels = c("Female", "Male"), guide = guide_legend(keyheight = unit(2.5, units = "mm"),  keywidth = unit(12, units = "mm"), label.position = "bottom", title.position = 'top', nrow = 1, reverse = TRUE)) +
  scale_x_continuous(breaks = c(-breaks, spacing + breaks), labels = c(breaks, breaks)) +
  theme_minimal() +
  labs(x = NULL, y = NULL, title = NULL, subtitle = NULL, caption = NULL, fill = NULL) +
  theme(plot.margin = unit(rep(0.5, 4), "cm"),
        panel.background = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "top", 
        legend.justification = "left",
        legend.direction = "horizontal",
        legend.text = element_text(size = 10),
        axis.text.y = element_blank(),
        axis.ticks = element_line(colour = NA))

div(class = "caption", p("Source: PHE Covid-19 Situational Awareness Explorer | ONS"))
```
</div>
</div>

<div class = "col-sm-6">
<div class = "div">
```{r, dev = 'svg'}
# Ethnicity
ethnic_group <- cases %>% 
  # last 7 days
  filter(date >= max(date)-days(6) & date <= max(date)) %>% 
  mutate(ethnic_group = 
           case_when(
             Ethnicity %in% c("British", "Irish", "Any other White background") ~ "White",
             Ethnicity %in% c("White and Black Caribbean", "White and Black African", "White and Asian", "Any other Mixed / Multiple ethnic background") ~ "Mixed",
             Ethnicity %in% c("Indian", "Pakistani", "Bangladeshi", "Any other Asian background") ~ "Asian / Asian British",
             Ethnicity %in% c("Caribbean", "African", "Any other Black / African / Caribbean background") ~ "Black / Black British",
             Ethnicity %in% c("Chinese", "Any other ethnic group") ~ "Other",
             TRUE ~ "Not known")) %>% 
  group_by(ethnic_group) %>% 
  summarise(n = n()) %>% 
  ungroup() %>% 
  filter(ethnic_group != "Not known")

ethnic_group_table <- left_join(ethnicity, ethnic_group, by = "ethnic_group") %>% 
  mutate(rate = round(n/population*100000,1)) %>% 
  select(-population) %>% 
  reactable( 
    resizable = TRUE,
    sortable = TRUE,
    defaultSorted = "rate",
    defaultSortOrder = "desc",
    defaultColDef = colDef(align = "left"),
    defaultColGroup = colGroup(align = "left"), 
    columns = list(
      ethnic_group = colDef(name = "Ethnic group"),
      n = colDef(name = "n", format = colFormat(separators = TRUE)),
      rate = colDef(name = "per 100,000", format = colFormat(separators = TRUE))
    ),
    columnGroups = list(
      colGroup(name = "Cases in last 7 days", columns = c("n", "rate"))
    )
  )

div(class = "div",
    div(class = "subtitle",
        div(class = "title", "Case count and incidence per 100,000 population by ethnicity"),
        paste("Trafford,", format(max(cases$date)-days(6), "%d %B"), "to", format(max(cases$date), "%d %B %Y"))
    ),
    br(),
    ethnic_group_table,
    div(class = "caption", p("Source: PHE Covid-19 Situational Awareness Explorer | ONS"))
)
```
</div>
</div>
</div>
</div>
<br /><br />

**Notes**

- Patients who self-identify as Chinese or as belonging to "Any other ethnic group" are classified as "Other." This follows the ethnicity classification used by Hospital Episode Statistics (see Appendix B of Public Health England's [Disparities in the risk and outcomes of COVID-19](https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/908434/Disparities_in_the_risk_and_outcomes_of_COVID_August_2020_update.pdf) report for a comparison with ONS classifications).

---


